---
output: rmarkdown::github_document
#bibliography: README-refs.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# R/`tmle3`

[![Travis-CI Build Status](https://travis-ci.org/jeremyrcoyle/tmle3.svg?branch=master)](https://travis-ci.org/jeremyrcoyle/tmle3)
[![AppVeyor Build  Status](https://ci.appveyor.com/api/projects/status/github/jeremyrcoyle/tmle3?branch=master&svg=true)](https://ci.appveyor.com/project/jeremyrcoyle/tmle3)
[![Coverage Status](https://img.shields.io/codecov/c/github/jeremyrcoyle/tmle3/master.svg)](https://codecov.io/github/jeremyrcoyle/tmle3?branch=master)
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](http://www.repostatus.org/badges/latest/wip.svg)](http://www.repostatus.org/#wip)
[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](http://www.gnu.org/licenses/gpl-3.0)

> _Extensible_ TMLE framework

__Author:__ [Jeremy Coyle](https://github.com/jeremyrcoyle)

---

## What's `tmle3`?

`tmle3` is a general framework that supports the implementation of a range of TMLE parameters with a unified interface. The goal is that the `tmle3` framework be as general as the mathematical framework upon which it's based. 

---

## Installation

You can install the development version of `tmle3` from GitHub via
[`devtools`](https://www.rstudio.com/products/rpackages/devtools/) with

```{r gh-installation, eval = FALSE}
devtools::install_github("jeremyrcoyle/tmle3")
```

---

## Getting Started

The best place to get started is the Framework Overview, which describes the individual components of the `tmle3` framework. You can find it here: https://tmle3.tlverse.org/articles/framework.html.

-- 

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/jeremyrcoyle/tmle3/issues).

---

## License

&copy; 2017 [Jeremy R. Coyle](https://github.com/jeremyrcoyle)

The contents of this repository are distributed under the GPL-3 license. See
file `LICENSE` for details.

