---
output:
  rmarkdown::github_document
bibliography: "README-refs.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# R/`tmle3`

[![Travis-CI Build Status](https://travis-ci.org/tlverse/tmle3.svg?branch=master)](https://travis-ci.org/tlverse/tmle3)
[![Appveyor Build Status](https://ci.appveyor.com/api/projects/status/cxp6a15anauyadgb?svg=true)](https://ci.appveyor.com/project/tlverse/tmle3)
[![Coverage Status](https://img.shields.io/codecov/c/github/tlverse/tmle3/master.svg)](https://codecov.io/github/tlverse/tmle3?branch=master)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](http://www.gnu.org/licenses/gpl-3.0)

> The _Extensible_ TMLE framework

__Author:__ [Jeremy Coyle](https://github.com/jeremyrcoyle)

---

## What's `tmle3`?

`tmle3` is a general framework that supports the implementation of a range of
Targeted Maximum Likelihood / Minimum Loss-Based Estimation (TMLE) parameters
through exposing a unified interface. The goal is that the `tmle3` framework
be as general as the mathematical framework upon which it's based.

For a general discussion of the framework of targeted minimum loss-based
estimation and the role this methodology plays in statistical and causal
inference, the canonical references are @vdl2011targeted and @vdl2018targeted.

---

## Installation

<!--
For standard use, we recommend installing the package from
[CRAN](https://CRAN.R-project.org/package=hal9001) via

```{r cran-installation, eval = FALSE}
install.packages("hal9001")
```
-->

To contribute, install the _development version_ of `tmle3` from GitHub via
[`remotes`](https://CRAN.R-project.org/package=remotes):

```{r gh-installation, eval = FALSE}
remotes::install_github("tlverse/tmle3")
```

---

## Getting Started

The best place to get started is the "Framework Overview" document, which
describes the individual components of the `tmle3` framework. It may be found at
https://tlverse.org/tmle3/articles/framework.html.

---

## Issues

If you encounter any bugs or have any specific feature requests, please [file an
issue](https://github.com/tlverse/tmle3/issues).

---

## License

&copy; 2017-2020 [Jeremy R. Coyle](https://github.com/jeremyrcoyle)

The contents of this repository are distributed under the GPL-3 license. See
file `LICENSE` for details.

---

## References

