---
title: "`tmle3` Framework Overview"
author: "[Jeremy Coyle](https://github.com/jeremyrcoyle)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tmle3 Framework Overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, echo=FALSE, warning=FALSE, eval=FALSE}
library(knitr)

# define insert_fun
insert_fun <- function(name) {
  knitr::read_chunk(lines = utils::capture.output(dump(name, "")),
                    labels = paste(name, "source", sep = "-"))
}
```

# Introduction

The `tmle3` package differs from previous TMLE software efforts in that it
attempts to directly model the key objects defined in the mathematical and
theoretical framework of Targeted Minimum Loss-Based Estimation (TMLE). That is,
rather than focus on implementing a specific TML estimator, or a small set of
related estimators, the focus is on modeling the TMLE _framework_ itself.

Therefore, we explicitly define objects to model the NPSEM, the factorized
likelihood, counterfactual interventions, parameters, and TMLE update
procedures. The hope is that, in so doing, it will be possible to support a
substantial subset of the vast array of TML estimators currently present in the
literature, as well as those that have yet to be developed. In this vignette, we
describe these mathematical objects, their software analogs in `tmle3`, and
illustrate with a motivating example, described below. At the end, we describe
how these objects can be bundled into a complete specification of a TML
estimation procedure that can be easily applied by an end user.

### Motivating Example

We use data from the Collaborative Perinatal Project (CPP), available in the
`sl3` package. To simplify this example, we define a binary intervention
variable, `parity01` -- an indicator of having one or more children before the
current child and a binary outcome, `haz01` -- an indicator of having an above
average height for age.

```{r prelims}
library(tmle3)
library(sl3)
data(cpp)
cpp <- cpp[!is.na(cpp[, "haz"]), ]
cpp$parity01 <- as.numeric(cpp$parity > 0)
cpp[is.na(cpp)] <- 0
cpp$haz01 <- as.numeric(cpp$haz > 0)
```

# NPSEM

TMLE requires the specification of a Nonparametric Structural Equation Model
(NPSEM), which specifies our knowledge of relationships between the variables.

We start with a set of endogenous variables, $X=(X_1,\ldots,X_J)$, that we want
to model the relationship between. Each $X_j$ is at least partially observed in
the dataset. The NPSEM defines each variable ($X_j$) by a deterministic function
($f_{X_j}$) of its parent nodes ($Pa(X_j)$) and an exogenous random variable
($U_{X_j}$):

$$X_j = f_{X_j}(Pa(X_j), U_{X_j}),\;\; j\in \{1, \ldots, J\}$$

The exact functional form of the functions $f_{X_j}$ is left unspecified at this
step. If there is _a priori_ knowledge for some of these functions, that can be
specified during the likelihood step below.

### Causal Considerations

The collection of exogenous random variables defined by the NPSEM is
$U = (U_{X_1}, \ldots, U_{X_J})$. Typically, non-testable assumptions about the
joint distribution of $U$ are necessary for identifiability of causal parameters
with statistical parameters of the observed data. These assumptions are not
managed in the `tmle3` framework, which instead focus on the statistical
estimation problem. Therefore, those developing tools for end users need to be
clear about the additional causal assumptions necessary for causal
interpretation of estimates.

### Example

In the case of our CPP example, we use the classic point treatment NPSEM which
defines three nodes: $X = (W, A, Y)$, where $W$ is a set of baseline covariates,
$A$ is our exposure of interest (`parity01`), and $Y$ is our outcome of interest
(`haz01`). We define the following SCM:

$$W = f_W(U_W)$$
$$A = f_A(W, U_A)$$
$$Y = f_Y(W, U_Y)$$

In `tmle3`, this is done using the `define_node` function for each node.
`define_node` allows a user to specify the node_name, which columns in the data
comprise the node, and a list of parent nodes.

``` {r define_npsem}
npsem <- list(
  define_node("W", c(
    "apgar1", "apgar5", "gagebrth", "mage",
    "meducyrs", "sexn"
  )),
  define_node("A", c("parity01"), c("W")),
  define_node("Y", c("haz01"), c("A", "W"))
)
```

Nodes also track information about the data types of the variables (continuous,
categorical, binomial, etc). Here, that information is being estimated
automatically from the data. In the future, each node will also contain
information about censoring indicators, where applicable, but this is not yet
implemented.

### `tmle3_Task`

A `tmle3_Task` is an object comprised of observed data, and the NPSEM defined
above:

```{r define_task}
tmle_task <- tmle3_Task$new(cpp, npsem = npsem)
```

This task object contains methods to help subset the data as needed for various
steps in the TMLE process:

```{r subset_task}
# get the outcome node data
head(tmle_task$get_tmle_node("Y"))

# get the sl3 task corresponding to an outcome regression
tmle_task$get_regression_task("Y")
```

A `tmle3_Task` is a special kind of `sl3_Task` that can be used to estimate
factors of a likelihood from data. The process of defining and estimating a
likelihood is described next.

# Likelihood

Having defined the NPSEM, we can now define a joint likelihood (probability
density function) over the observed variables $X$:

$$P(X_1, \ldots, X_J \in D) = \int_D f_{X_1, \ldots, X_J}(x_1, \ldots, x_J)
dx_1, \ldots, dx_J$$

This can then be factorized into a series of conditional densities according to
the NPSEM:
$$f_{X_1, \ldots, X_J} = \prod_j^J f_{X_j \mid Pa(X_j)}(x \mid Pa(x_j))$$

Where each $f_{X_j \mid Pa(X_j)}$ is a conditional pdf (or probability mass
function for discrete $X_j$), where the conditioning set is all parent nodes as
defined in the NPSEM. We refer to these objects as _likelihood factors_.

TMLE depends on estimates (or _a priori_ knowledge) of the functional form of
these likelihood factors. However, not all factors of the likelihood are always
necessary for estimation, and only those necessary will be estimated.

### Likelihood Factor Objects

`tmle3` models this likelihood as a list of likelihood factor objects, where
each likelihood factor object describes either _a priori_ knowledge or an
estimation strategy for the corresponding likelihood factor. These objects all
inherit from the `LF_base` base class, and there are different types depending
on which of a range of estimation strategies or _a priori_ knowledge is
appropriate.

In some cases, a full conditional density for a particular factor is not
necessary. Instead, a conditional mean -- a much easier quantity to estimate --
is all that's required. Although conditional means are not truly likelihood
factors, conditional means are also modeled using using likelihood factor
objects.

### `LF_emp`

`LF_emp` represents a likelihood factor to be estimated using nonparametric
maximum likelihood estimation (NP-MLE). That is, probability mass $\frac{1}{n}$
is placed on each observation in the observed dataset:

$$f_{X_j}(x_j) = \frac{1}{n}\mathbb{I}(x_j \in X_{n,j})$$

Going forward, weights will be used if specified, although this is not yet
supported. `LF_emp` only supports marginal densities. That is, the conditioning
set, $Pa(X_j)$ must be empty. Therefore, it is only appropriate for estimation
of the marginal density of baseline covariates.

### `LF_fit`

`LF_fit` represents a likelihood factor to be estimated using the `sl3`
framework. Based on the learner type used, this can fit a pmf (for binomial or
categorical data, see `sl3_list_learners("binomial")` and
`sl3_list_learners("categorical")` for lists), a conditional mean (most
learners), or a conditional density (e.g., using a semiparametric conditional
density estimator via `Lrnr_density_semiparametric`).  `LF_fit` takes a `sl3`
learner object as an argument, which is fit to the data in the `tmle3_Task`
automatically. Details for specifying different kinds of learners in `sl3` may
be found at http://tlverse.org/sl3/articles/intro_sl3.html

### Specifying _a priori_ knowledge.

The above to likelihood factor types, `LF_fit`, and `LF_emp`, are both
likelihood factors where the factor is estimated from data. In some cases, users
may have _a priori_ knowledge of a likelihood factor. For instance, in an RCT,
there might be an unconditional probability of treatment of $p = 0.5$.
Additional likelihood factor types need to be create to accommodate this type of
knowledge.

### Example

Going back to our CPP data example, we will estimate the marginal likelihood of
$W$, using NP-MLE, the conditional density of $A$ given $W$ using a GLM fit via
`sl3` and the conditional mean of $Y$ given $A$ and $W$ using another GLM fit
via `sl3`:

```{r likelihood}
# set up sl3 learners for tmle3 fit
lrnr_glm_fast <- make_learner(Lrnr_glm_fast)
lrnr_mean <- make_learner(Lrnr_mean)

# define and fit likelihood
factor_list <- list(
  define_lf(LF_emp, "W"),
  define_lf(LF_fit, "A", lrnr_glm_fast),
  define_lf(LF_fit, "Y", lrnr_glm_fast, type="mean")
)
```

The particular likelihood factors and estimation strategies to use will of
course depend on the parameter of interest. Once this list of likelihood factors
is defined, we can construct a `Likelihood` object and train it on the data
contained in `tmle_task`:

```{r likelihood_fit}
likelihood_def <- Likelihood$new(factor_list)
likelihood <- likelihood_def$train(tmle_task)
print(likelihood)
```

A `tmle3` `Likelihood` is actually a special type of `sl3` learner, so the
syntax to train it on data is analogous.

Having fit the likelihood, we can now get likelihood values for any
`tmle3_Task`:

```{r likelihood_values}
likelihood_values <- likelihood$get_likelihoods(tmle_task,"Y")
head(likelihood_values)
```

## Counterfactual Likelihoods

In `tmle3`, interventions are modeled by likelihoods where one or more
likelihood factors is replaced with a counterfactual version representing some
intervention.

`tmle3` defines the `CF_Likelihood` class, which inherits from `Likelihood`, and
takes an `observed_likelihood` and an `intervention_list`.

Below, we describe some examples of additional likelihood factors intended to be
used to describe interventions. We expect this list to grow as `tmle3` is
extended to additional use-cases.

### `LF_static`

Likelihood factor for a static intervention, where all observations are set do a
single intervention value $x'$:

$$f_{X_j \mid Pa(X_j)}(x_j \mid Pa(x_j)) = \mathbf{I}(x_j = x')$$

### Other intervention likelihood factor types

Additional likelihood factor types need to be defined for other types of
interventions, such as dynamic rules and stochastic interventions. Currently, a
prototype version of a stochastic shift intervention exists in `LF_shift`.

### Example

For our CPP example, we'll define a simple intervention where we set all
treatment $A = 1$:

```{r intervention}
intervention <- define_lf(LF_static, "A", value = 1)
```

We can then use this to construct a counterfactual likelihood:

```{r cf_likelihood}
cf_likelihood <- make_CF_Likelihood(likelihood, intervention)
```

A `cf_likelihood` is a likelihood object, and so has the same behavior as the
observed likelihood object defined above, but with the observed likelihood
factors being replaced by the defined intervention likelihood factors.

In particular, we can get likelihood values under the counterfactual likelihood:

```{r cf_likelihood_values}
cf_likelihood_values <- cf_likelihood$get_likelihoods(tmle_task, "A")
head(cf_likelihood_values)
```

We see that the likelihood values for the $A$ node are all either 0 or 1, as
would be expected from an indicator likelihood function. In addition, the
likelihood values for the non-intervention nodes have not changed.

### Counterfactual Tasks

Each `CF_Likelihood` can generate one or more counterfactual tasks. These are
`tmle3_Task`s in which observed values are replaced with counterfactual values
according to the specified intervention distribution. For deterministic
interventions, only one task will be generated. However, stochastic
interventions, when implemented, will generate several such tasks, one for each
combination of possible values of the intervention node(s).

To enumerate these tasks, use `enumerate_cf_tasks`:

```{r cf_likelihood_tasks}
cf_likelihood_tasks <- cf_likelihood$enumerate_cf_tasks(tmle_task)
head(cf_likelihood_tasks[[1]]$data)
```

In this case, you can see that `parity01` has been set to 1 for all
observations, consistent with a static intervention on this node.

# Update Procedure

In the TMLE framework, we define a target parameter $\Psi(P)$ as a mapping from
a probability distribution $P \in \mathcal{M}$ to a set of real numbers
$\mathbb{R}^d$. Here $\mathcal{M}$ is implied by the NPSEM we defined above.

In `tmle3`, we define parameter objects as objects inheriting from the
`Param_base` class, which keep track of not only the mapping from a probability
distribution to a parameter value, but also the corresponding EIF of the
parameter, and the "clever covariates" needed to calculate a TMLE update to the
likelihood.

Here, we define a treatment-specific mean (TSM) parameter based on the
intervention we defined previously:

```{r define_param}
tsm <- define_param(Param_TSM, likelihood, intervention)
```

**TODO**: provide details about parameter definition

# Update Procedure

The update procedure component of `tmle3` is currently in flux. The current
structure is as follows:

We have an object, `tmle3_Update`, which calculates the individual update steps
using `tmle3_Update$update_step`. This adds to a `Likelihood$update_list`, so
that future calls to `Likelihood$get_likelihoods` will return updated likelihood
values. However, likelihood values are generally recomputed at each step, which
requires applying all past updates. This is ridiculously inefficient.

Instead, we need to do what previous TMLE implementations have done, which is
enumerate a list of required likelihood values, and update those values as we go
(as opposed to updating the function and recalculating the value each time they
are needed). This requires the ability to have the parameters enumerate which
likelihood values they will need for defining the clever covariate, as well as
parameter mapping and the EIF. This has not yet been implemented.

Therefore, the update procedure, as well as the structure of the `Param_base`
parameter objects are subject to substantial changes in the near future.

Currently, the `tmle3_Update` object also has a hard-coded submodel (logistic),
loss function (log-likelihood), and solver (GLM). These need to be generalized
so updates can be done for a range of submodels, loss functions, and solvers.

Current Usage:
```{r update}
updater <- tmle3_Update$new()
targeted_likelihood <- Targeted_Likelihood$new(likelihood, updater)
```

# Target Parameter

In the TMLE framework, we define a target parameter $\Psi(P)$ as a mapping from
a probability distribution $P \in \mathcal{M}$ to a set of real numbers
$\mathbb{R}^d$. Here $\mathcal{M}$ is implied by the NPSEM we defined above.

In `tmle3`, we define parameter objects as objects inheriting from the
`Param_base` class, which keep track of not only the mapping from a probability
distribution to a parameter value, but also the corresponding EIF of the
parameter, and the "clever covariates" needed to calculate a TMLE update to the
likelihood.

Here, we define a treatment specific mean (TSM) parameter based on the
intervention we defined previously:

```{r param_tsm}
tsm <- define_param(Param_TSM, likelihood, intervention)
updater$tmle_params <- tsm
```

**TODO: provide details about parameter definition**



# `tmle3_Fit` - Putting it all together

Now that we have specified all the components required for the TMLE procedure,
we can generate an object that manages all the components and finally calculate
an appropriate TML estimator.

```{r fit_tmle}
tmle_fit <- fit_tmle3(tmle_task, targeted_likelihood, tsm, updater)
print(tmle_fit)
```

# TMLE Specification

The `tmle3` framework described above is completely general, and allows most
components of the TMLE procedure to be specified in a modular way. However, most
end users will not be interested in manually specifying all of these components.
Therefore, `tmle3` implements a `tmle3_Spec` object that bundles a set of
components into a _specification_ that, with minimal additional detail, can be
run by an end-user:

```{r tmle_spec}
nodes <- list(W = c("apgar1", "apgar5", "gagebrth", "mage", "meducyrs",
                    "sexn"),
              A = "parity01",
              Y = "haz01")

lrnr_glm_fast <- make_learner(Lrnr_glm_fast)
lrnr_mean <- make_learner(Lrnr_mean)
learner_list <- list(Y = lrnr_mean, A = lrnr_glm_fast)

# make a new copy to deal with data.table weirdness
cpp2 <- data.table::copy(cpp)

tmle_fit_from_spec <- tmle3(tmle_TSM_all(), cpp2, nodes, learner_list)
print(tmle_fit_from_spec)
```

Currently, this is effectively a hard-coded list of those details: the structure
of the NPSEM, the parameters, and the update procedure are coded into the
specification. Only the data, the roles of the variables, and the  `sl3`
learners to use for likelihood estimation. Ideally, instead a `tmle3_Spec` would
represent a set of reasonable defaults for a particular TMLE, that experienced
users could override where appropriate.

# Conclusion

Obviously, there's a lot more to do:

* Generalize `tmle3_Update`
* Generalize `tmpe3_Spec`
* Better handling of bounded continuous outcomes
* Expand documentation of parameter definitions
* Add support for dynamic rules and stochastic interventions
* CV-TMLE
* C-TMLE
* IPCW-TMLE
* Extension to longitudinal data settings
